Chemistry of the Orphan Stream I: Identifying Stream Members \\ from Low-Resolution Spectroscopy

### 2012-11-15 2012-oss-mrs/ms-aastex.tex
2012 mrs paper sync

\documentclass[10pt,apjl]{emulateapj}
\usepackage{amssymb,graphics} % for resolution symbol

@@

\shorttitle{Identifying Orphan Stream Members from Low-Resolution Spectroscopy}
\shortauthors{Casey et al}

\title{Hunting the Parent of the Orphan Stream: Identifying Stream Members \\ from Low-Resolution Spectroscopy}

@@

The Milky Way stellar halo has partly formed through the accretion of satellites which are disrupted by tidal forces as they fall into the Milky Way's potential. Stars which were once gravitationally bound to the satellite are strewn along the progenitor's orbit in leading and trailing streams of stars. The kinematics of the stream are sensitive to the shape of the galactic potential, allowing us to constrain the Milky Way potential, and reconstruct the formation history of the Galaxy. The level of accreted substructure in the Milky Way has only recently become apparent through multi-band photometric surveys like the Sloan Digital Sky Survey (SDSS). The more prominent of the detectable substructures, like Sagittarius, have been well-studied. One of the more prominent \--- yet less studied \--- substructures is that of the Orphan Stream.

The Orphan Stream was independently detected by both \citet{Grillmair_2006} and \citet{Belokurov_et-al_2006}, and is distinct from other substructures in the halo. The stream stretches over $60\,^\circ$ in the sky, has a low surface brightness and a narrow stream width of only $\sim$2\,$^\circ$ and as the name suggests, the parent object largely remains a mystery. The stream extends past the celestial equator \--- outside the SDSS footprint \--- but has not been detected in existing southern surveys \citep{Newberg_et-al_2010}. Whilst the parent system remains elusive, significant effort has been placed on associating the stream with known Milky Way satellites \citep{Zucker_et-al_2006, Fellhaur_et-al_2007,Jin_Lynden_Bell_2007,Sales_et-al_2008}. In contrast, there has been relatively limited observational work on the Orphan Stream itself other than the original discovery papers \citep{Grillmair_Dionatos_2006, Belokurov_et-al_2006, Belokurov_et-al_2007} and the work of \citet{Newberg_et-al_2010}. This is largely to be expected given the absence of multi-band photometry in the southern sky and the low total luminosity of the stream, making it difficult to reliably separate Orphan Stream members from halo stars. Further understanding on the extent of the stream awaits the SkyMapper and Pan-STARRS photometric surveys \citep{Keller_et-al_2007, Hodapp_et-al_2004}.

As \citet{Sales_et-al_2008} point out, there is a natural observational bias towards more massive and recent mergers like Sagittarius. Consequently, the fainter end of this substructure distribution has yet to be fully recovered, or thoroughly examined. Interestingly, there are indications that  some fainter substructures like the Orphan Stream and the Palomar 5 tidal tails \citep{Odenkirchen_et-al_2009} have orbits which seem to be best-fit by Milky Way models with nearly 60\% less mass \citep{Newberg_et-al_2010} than generally reported by \citet{Xue_et-al_2008} and \citet{Koposov_et-al_2010}. Such a discrepancy in the mass of the Milky Way is troublesome. More complete photometric and kinematic maps of these low total luminosity streams may provide the best test as to whether this mass discrepancy is real, or an artefact of incomplete observations. Whilst the full spatial extent of the Orphan Stream remains incomplete, we can examine the detailed chemistry of its members and investigate the stream history, as well as make predictions about the nature of the progenitor.

@@

Given our colour selection we expect to recover giants and contaminating dwarf contaminants. Although the 2MASS $JHK$ colours can help to separate dwarfs and giants, our target K-giants stars are too faint to be detected in the 2MASS catalogue.

@@

Observations took place on the Anglo-Australian Telescope using the AAOmega spectrograph in April 2009. AAOmega is a fibre-fed, dual beam multi-object spectrograph which is capable of simultaneously observing spectra of 392 (science and sky) targets across a $2\,^\circ$ field of view. We used the 5700\,{\AA} dichroic in combination with the 1000I grating in the red arm, and the 580V grating in the blue arm. This provides a spectral coverage between $800 \leq \lambda \leq 950$\,nm in the red at $\mathcal{R} \approx 4400$, and between $370 \leq \lambda \leq 580$\,nm with a lower spectral resolution of $\mathcal{R} \approx 1300$ in the blue.

The data were reduced using the standard \textsc{2DFDR} reduction pipeline\footnote{http://www.aao.gov.au/2df/aaomega/aaomega\_2dfdr.html}. After flat-fielding, throughput calibration for each fibre was achieved using the intensity of skylines in each fibre. The median flux of dedicated sky fibres was taken for sky subtraction, and wavelength calibration was performed using arc lamp exposures between science frames. Three thirty minute science exposures were median-combined to assist with cosmic ray removal. The median S/N obtained in the red arm for our fields is modest at 35 per pixel. With the presence of strong Ca\,\textsc{II} triplet lines in the red arm we are able to ascertain reliable radial velocities and reasonable estimates on overall metallicity \citep[][and references therein]{Starkenburg_et-al_2010}. Our spectral region also includes gravity-sensitive magnesium lines: Mg\,\textsc{I} at 8807\,{\AA}, and the Mg\,\textsc{I}\,b 3$p$-4$s$ triplet lines at $\sim$5178\,{\AA}. As we demonstrate in the next section, these lines are sufficient to discriminate dwarfs from giants even with weak signal.

@@

We have employed a combination of separate criteria to identify likely Orphan Stream members: kinematics, a giant/dwarf indication from Mg\,\textsc{I} lines, and consistent metallicities from isochrone fitting and the Ca\,\textsc{II} triplet lines. Each criteria is discussed here separately.

@@

Figure \ref{fig:velocities} shows a histogram of our galactocentric velocities, compared to the predicted smooth line-of-sight velocity distribution for this region from the Besan\c{c}on model \citep{Robin_et-al_2003}. We have selected predicted stars from the Besan\c{c}on model using the same criteria as outlined in \S\ref{sec:target-selection} after using the \citet{Jordi_et-al_2006} colour transformations. It is clear that our target selection has yielded mostly nearby thick disk dwarf stars with $V_{GSR} \approx -120$\,km s$^{-1}$. The expected kinematic signature for the Orphan Stream, as \citet{Newberg_et-al_2010} reported, is labelled in Figure \ref{fig:velocities}. There is no obvious sharp kinematic peak representative of the Orphan Stream in our sample. From kinematics alone, our targets appears largely indistinguishable from a smooth halo distribution.

@@

We have measured the equivalent width of the gravity-sensitive Mg\,\textsc{I} line at 8807 \AA{} to distinguish dwarfs from giants \citep{Battaglia_Starkenburg_2012}. At a given temperature (or $g - r$) and metallicity, giant stars present narrower Mg\,\textsc{I} absorption lines than their dwarf contaminants. Given our target selection, our sample is likely to contain many more dwarfs than giants. In some cases no Mg\,\textsc{I} 8807 \AA{} line was apparent, so an upper limit was estimated based on the S/N. In these cases the candidate was considered a ``non-dwarf" because we cannot exclusively rule out a metal-poor sub-giant with this criteria alone. For these purposes we are only looking for a simple indication as to whether a star is likely a dwarf or not. Figure \ref{fig:ew-mg} illustrates the trend with EW$_{\lambda8807}$ against SDSS de-reddened\footnote{All magnitudes presented in this letter are de-reddened using the \citet{Schlegel_Finkbeiner_Davis_1998} dust maps.} $g - r$, illustrating the dominant upper dwarf branch we wish to exclude. The adopted dwarf/giant separation line is shown. Admittedly this selection criteria is not particularly strict, but we are allowing multiple inclusive cuts and taking the intersection of these criteria to identify Orphan Stream giants.

@@

Our dwarf/giant separation line in Figure \ref{fig:ew-mg} yields 425 potential giants. This fraction of giants ($\sim$44\%) is much larger than that found by \citeauthor{Casey_et-al_2012} \citeyear{Casey_et-al_2012}a who employed a similar colour selection. Hence, our gravity criteria is extremely inclusive and susceptible to dwarf contamination. Upon taking the intersection of our kinematic and gravity selections, we deduce eighteen stars that appear likely Orphan Stream K giants.

@@

We have measured the equivalent widths of the Ca\,\textsc{II} near-infrared triplet lines for stars that meet our kinematic and surface gravity criteria. After correcting for luminosity, the equivalent width of the Ca\,\textsc{II} triplet lines provide a good indication of the overall metallicity of a RGB star \citep{Amandroff_Da_Costa_1991}. We have employed the \citet{Starkenburg_et-al_2010} relationship and corrected for luminosity in $g$ against the horizontal branch magnitude at $g_{HB}$ = 17.1 \citep{Newberg_et-al_2010}. Strictly speaking, the Ca\,\textsc{II}\---[Fe/H] calibration is only valid for stars brighter than the horizontal branch, although the relationship only becomes significantly inappropriate near $g \-- g_{HB} \sim +1$ \citep{Saviane_et-al_2012}. Many of our candidates are fainter than this valid luminosity range, and therefore they should not be excluded solely because of their derived metallicities. Stars fainter than $g_{HB}$ will have slightly lower metallicities than predicted by our Ca\,\textsc{II}\---[Fe/H] relationship, and for these stars we will only use metallicities to assign a relative qualitative likelihood for stream membership.
Given a distance to the Orphan Stream, we can also estimate a star's metallicity through isochrone fitting. We have used a 10\,Gyr \citet{Girardi_et-al_2008} isochrone at 21.4\,kpc \citep{Newberg_et-al_2010} and found metallicities for all eighteen likely stream members from their best-fitting isochrone. Derived metallicities from Ca\,\textsc{II} line strengths and isochrone fitting that are consistent (within $\pm0.3$\,dex) indicates these measurements are reliable, and that these stars are indeed at a distance of $\sim$21.4\,kpc. We find ten highly likely stream members with consistently derived metallicities. They fall within the shaded region illustrated in Figure \ref{fig:feh}.

@@

Given that our Orphan Stream giants cover a wide evolutionary range along the giant branch (Figure \ref{fig:cmd}), we are in a good position to revise the distance estimate to the stream. Given a 10\,Gyr \citet{Girardi_et-al_2008} isochrone at $[\mbox{Fe/H}] = -1.63$, we find a best-fitting distance to the stream of $22.5 \pm 1.0$\,kpc at $(l, b) = (250\,^\circ,\,50\,^\circ)$. This isochrone is shown in Figure \ref{fig:cmd}, illustrating that three of our nine most probable member stars lie perfectly on this isochrone. Our derived distance is in reasonably good agreement with the $21.4 \pm 1.0$\,kpc independently estimated by \citet{Grillmair_2006} and \citet{Newberg_et-al_2010}.

@@

% Adding \newpage here because otherwise only the Conclusions header is shown at the end of Page 4.
\newpage

@@

We have presented a detailed analysis to isolate individual Orphan Stream K giants from low-resolution spectroscopy using a combination of photometric, kinematic, gravity, metallicity, and proper motion information. Although each individual criteria is likely to induce some level contamination, their intersection reveals nine highly probable, self-consistent, Orphan Stream K giants.  We deduce a median stream metallicity of $[\mbox{Fe/H}] = -1.63 \pm 0.19$ and find an intrinsically wide metallicity spread of $\sigma = 0.56$\,dex, indicative of a dSph origin. Unlike other stellar tracers, K-type giants can exist at all metallicities, hence our derived metallicity spread is likely representative of the true stream metallicity distribution function. Recall that the metallicity determination was performed after kinematic and gravity cuts, and three of our most probable members lay perfectly on a 10\,Gyr isochrone of $[\mbox{Fe/H}] = -1.63$. Our stream members indicate a distance to the stream of $22.5 \pm 1.0$\,kpc at $(l, b) = (250\,^\circ,\,50\,^\circ)$, in agreement with that deduced by \citet{Grillmair;et-al_2006} and \citet{Newberg_et-al_2010}.

@@

The K giants presented here can provide great insight into the chemistry, and history of the Orphan Stream. High-resolution spectroscopic observations have been taken for some of our most highly probable members and a detailed chemical analysis will be presented in a forthcoming paper (Casey et al. 2012c, in preparation). Detailed chemical abundances can help determine both the nature of the progenitor before it's discovered, and allows us to compare peculiar chemical signatures with those of the known Milky Way satellites in order to associate likely parents. However, at least for the moment, the Orphan Stream remains appropriately named.

@@

\bibitem[{{Grillmair} (2006)}]{Grillmair_2006}
{Grillmair}, C.~J. 2006, \apj, 643, L37

>>>>

\documentclass[10pt,apjl]{emulateapj}
\usepackage{amssymb} % for resolution symbol

@@

\shorttitle{Chemistry of the Orphan Stream I}
\shortauthors{Casey et al}

\title{Chemistry of the Orphan Stream I: Identifying Stream Members \\ from Low-Resolution Spectroscopy}

@@

The Milky Way stellar halo has partly formed through the accretion of satellites which are disrupted by tidal forces as they fall into the Milky Way's potential. Stars which were once gravitationally bound to the satellite are left behind in a stream of stars. Stream kinematics are sensitive to the shape of the galactic potential, allowing us to constrain the Milky Way potential, and reconstruct the formation history of the Galaxy. The level of accreted substructure in the Milky Way has only recently become apparent through multi-band photometric surveys like the Sloan Digital Sky Survey (SDSS). The more prominent of the detectable substructures, like Sagittarius, have been well-studied. One of the more prominent \--- yet less studied \--- substructures is that of the Orphan Stream.

The Orphan Stream was independently detected by both \citet{Grillmair_Dionatos_2006} and \citet{Belokurov_et-al_2006}, and is distinct from other substructures in the halo. The stream stretches over $60\,^\circ$ in the sky, has a low surface brightness and a narrow stream width of only $\sim$2\,$^\circ$ and as the name suggests, the parent object largely remains a mystery. The stream extends past the celestial equator \--- outside the SDSS footprint \--- but has not been detected in existing southern surveys \citep{Newberg_et-al_2010}. Whilst the parent system remains elusive, significant effort has been placed on associating the stream with known Milky Way satellites \citep{Zucker_et-al_2006, Fellhaur_et-al_2007,Jin_Lynden_Bell_2007,Sales_et-al_2008}. In contrast, there has been relatively limited observational work on the Orphan Stream itself other than the original discovery papers \citep{Grillmair_Dionatos_2006, Belokurov_et-al_2006, Belokurov_et-al_2007} and the work of \citet{Newberg_et-al_2010}. This is largely to be expected given the absence of multi-band photometry in the southern sky and the low total luminosity of the stream, making it difficult to reliably separate Orphan Stream members from halo stars. Further understanding on the extent of the stream awaits the SkyMapper and Pan-STARRS photometric surveys \citep{Keller_et-al_2007, Hodapp_et-al_2004}.

As \citet{Sales_et-al_2008} point out, there is a natural observational bias towards more massive and recent mergers like Sagittarius. Consequently, the fainter end of this substructure distribution has yet to be fully recovered, or thoroughly examined. Interestingly, there are indications that  some fainter substructures like the Orphan Stream and the Palomar 5 tidal tails \citep{Odenkirchen_et-al_2009} have orbits which seem to be best-fit by Milky Way models with nearly 60\% less mass \citep{Newberg_et-al_2010} than generally reported by \citet{Xue_et-al_2008} and \citet{Koposov_et-al_2010}. Such a discrepancy in the mass of the Milky Way is troublesome. More complete photometric and kinematic maps of these low total luminosity streams may provide the best test as to whether this mass discrepancy is real, or an artefact of incomplete observations. Whilst the full spatial extent of the Orphan Stream remains incomplete, we can examine the detailed chemistry of its members and investigate the stream history, as well as make predictions about the nature of the host system.

@@

Given our colour selection we expect to recover giants, and dwarf contaminants. The lack of 2MASS photometry for the selected stars means we cannot use the $(J-H, H-K)$ diagram to aid the separation of candidate giants from contaminating dwarfs.

@@

Observations took place on the Anglo-Australian Telescope using the AAOmega spectrograph in April 2009. AAOmega is a fibre-fed, dual beam multi-object spectrograph which is capable of simultaneously observing spectra of 392 (science and sky) targets across a $2\,^\circ$ field of view. We used the 5700\,{\AA} dichroic in combination with the 1000I grating in the red arm, and the 580V grating in the blue arm. This provides a spectral coverage between $800 \leq \lambda \leq 950$\,nm in the red at $\mathcal{R} \approx 4400$, and between $370 \leq \lambda \leq 580$\,nm with a lower resolving power of $\mathcal{R} \approx 1300$ in the blue.

The data were reduced using the standard \textsc{2DFDR} reduction pipeline\footnote{http://www.aao.gov.au/2df/aaomega/aaomega\_2dfdr.html}. After flat-fielding, throughput calibration for each fibre was achieved using the intensity of skylines in each fibre. The median flux of dedicated sky fibres was taken for sky subtraction, and wavelength calibration was performed using arc lamp exposures between science frames. Three thirty minute science exposures were median combined to assist with cosmic ray removal. The median S/N obtained in the red arm for our fields is modest at 35 per pixel. With the presence of strong Ca\,\textsc{II} triplet lines in the red arm we are able to ascertain reliable radial velocities and reasonable estimates on overall metallicity \citep[][and references therein]{Starkenburg_et-al_2010}. Our spectral region also includes gravity-sensitive magnesium lines: Mg\,\textsc{I} at 8807\,{\AA}, and the Mg\,\textsc{I}\,b 3$p$-4$s$ triplet lines at $\sim$5178\,{\AA}. As we demonstrate in the next section, these lines are sufficient to discriminate dwarfs from giants even with weak signal.

@@

We have employed a combination of separate criterion to identify likely Orphan Stream members: kinematics, a giant/dwarf indication from Mg\,\textsc{I} lines, and consistent metallicities from isochrone fitting and the Ca\,\textsc{II} triplet lines. Each criterion is discussed here separately.

@@

Figure \ref{fig:velocities} shows a histogram of our galactocentric velocities, compared to the predicted smooth line-of-sight velocity distribution for this region from the Besan\c{c}on model \citep{Robin_et-al_2003}. We have selected predicted stars from the Besan\c{c}on model using the same criterion as outlined in \S\ref{sec:target-selection} after using the \citet{Jordi_et-al_2006} colour transformations. It is clear that our target selection has yielded mostly nearby thick disk dwarf stars with $V_{GSR} \approx -120$\,km s$^{-1}$. The expected kinematic signature for the Orphan Stream, as \citet{Newberg_et-al_2010} reported, is labelled in Figure \ref{fig:velocities}. There is no obvious sharp kinematic peak representative of the Orphan Stream in our sample. From kinematics alone, our targets appears largely indistinguishable from a smooth halo distribution.

@@

We have measured the equivalent width of the gravity-sensitive Mg\,\textsc{I} line at 8807 \AA{} to distinguish dwarfs from giants \citep{Battaglia_Starkenburg_2012}. At a given temperature (or $g - r$) and metallicity, giant stars present narrower Mg\,\textsc{I} absorption lines than their dwarf contaminants. Given our target selection, our sample is likely to contain many more dwarfs than giants. In some cases no Mg\,\textsc{I} 8807 \AA{} line was apparent, so an upper limit was estimated based on the S/N. In these cases the candidate was considered a ``non-dwarf" because we cannot exclusively rule out a metal-poor sub-giant with this criterion alone. For these purposes we are only looking for a simple indication as to whether a star is likely a dwarf or not. Figure \ref{fig:ew-mg} illustrates the trend with EW$_{\lambda8807}$ against SDSS de-reddened\footnote{All magnitudes presented in this letter are de-reddened using the \citet{Schlegel_Finkbeiner_Davis_1998} dust maps.} $g - r$, illustrating the dominant upper dwarf branch we wish to exclude. The adopted dwarf/giant separation line is shown. Admittedly this selection criterion is not particularly strict, but we are allowing multiple inclusive cuts and taking the intersection of these criteria to identify Orphan Stream giants.

@@

Our dwarf/giant separation line in Figure \ref{fig:ew-mg} yields 425 potential giants. This fraction of giants ($\sim$44\%) is much larger than what would be expected given our selection criterion (e.g. see \citeauthor{Casey_et-al_2012} \citeyear{Casey_et-al_2012}a). Hence, our gravity criterion is extremely inclusive and susceptible to dwarf contamination. Upon taking the intersection of our kinematic and gravity selections, we deduce eighteen stars that appear likely Orphan Stream K giants.

@@

We have measured the equivalent widths of the Ca\,\textsc{II} near-infrared triplet lines for stars that meet our kinematic and surface gravity criteria. After correcting for luminosity, the equivalent width of the Ca\,\textsc{II} triplet lines provide a good indication of the overall metallicity of a RGB star \citep{Amandroff_Da_Costa_1991}. We have employed the \citet{Starkenburg_et-al_2010} relationship and corrected for luminosity in $g$ against the horizontal branch magnitude at $g_{HB}$ = 17.1 \citep{Newberg_et-al_2010}. Strictly speaking, the Ca\,\textsc{II}\---[Fe/H] calibration is only valid for stars brighter than the horizontal branch, although the relationship only becomes significantly inappropriate near $g \-- g_{HB} \sim +1$ \citep{Saviane_et-al_2012}. Many of our candidates are fainter than this valid luminosity range, and therefore they should not be excluded solely because of their derived metallicities. Stars fainted than $g_{HB}$ will have slightly lower metallicities than predicted by our Ca\,\textsc{II}\---[Fe/H] relationship, and for these stars we will only use metallicities to assign a relative qualitative likelihood for stream membership.
Given a distance to the Orphan Stream, we can also estimate a star's metallicity through isochrone fitting. We have used a 10\,Gyr \citet{Girardi_et-al_2008} isochrone at 21.4\,kpc \citep{Newberg_et-al_2010} and found metallicities for all eighteen likely stream members from their best-fitting isochrone. Derived metallicities from Ca\,\textsc{II} line strengths and isochrone fitting that are consistent (within $\pm0.3$\,dex) indicates these measurements are reliable, and that these stars are indeed at a distance of $\sim$21.4\,kpc. We find ten highly likely stream members with consistently derived metallicities. They fall within the shaded region illustrated in Figure \ref{fig:feh}.

@@

@@

Given that our Orphan Stream giants cover a wide evolutionary range along the giant branch (Figure \ref{fig:cmd}), we are in a good position to revise the distance estimate to the stream. Given a 10\,Gyr \citet{Girardi_et-al_2008} isochrone at $[\mbox{Fe/H}] = -1.63$, we find a best-fitting distance to the stream of $22.5 \pm 1.0$\,kpc at $(l, b) = (250\,^\circ,\,50\,^\circ)$. This isochrone is shown in Figure \ref{fig:cmd}, illustrating that three of our nine most probable member stars lie perfectly on this isochrone. Our derived distance is in reasonably good agreement with the $21.4 \pm 1.0$\,kpc deduced by \citet{Newberg_et-al_2010}.

@@

We have presented a detailed analysis to isolate individual Orphan Stream K giants from low-resolution spectroscopy using a combination of photometric, kinematic, gravity, metallicity, and proper motion information. Although each individual criterion is likely to induce some level contamination, their intersection reveals nine highly probable, self-consistent, Orphan Stream K giants.  We deduce a median stream metallicity of $[\mbox{Fe/H}] = -1.63 \pm 0.19$ and find an intrinsically wide metallicity spread of $\sigma = 0.56$\,dex, indicative of a dSph origin. Unlike other stellar tracers, K-type giants can exist at all metallicities, hence our derived metallicity spread is likely representative of the true stream metallicity distribution function. Recall that the metallicity determination was performed after kinematic and gravity cuts, and three of our most probable members lay perfectly on a 10\,Gyr isochrone of $[\mbox{Fe/H}] = -1.63$. Our stream members indicate a distance to the stream of $22.5 \pm 1.0$\,kpc at $(l, b) = (250\,^\circ,\,50\,^\circ)$, in agreement with that deduced by \citet{Newberg_et-al_2010}.

@@

The K giants presented here can provide great insight into the chemistry, and history of the Orphan Stream. High-resolution spectroscopic observations have been taken for some of our most highly probable members and a detailed chemical analysis will be presented in a forthcoming paper (Casey et al. 2012c, in preparation). Detailed chemical abundances can help determine both the nature of the host system before it's discovered, and allows us to compare peculiar chemical signatures with those of the known Milky Way satellites in order to associate likely parents. However, at least for the moment, the Orphan Stream remains appropriately named.

@@
